---
title: "GAE7"
author: "Mitch Mullin"
format: 
  html:
    theme: custom.scss
    embed-resources: true
    code-tools: true

csl: ecology.csl

editor: visual
bibliography: references.bib
---

```{r}
#| label: SetUp
#| message: FALSE
#| echo: FALSE

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
  ) 

library(emmeans)
library(lmerTest)
library(effectsize)
library(car)
library(ggfortify)
library(patchwork)
library(broom)
library(latex2exp)       
library(kableExtra)
library(tidyverse)

options(knitr.kable.NA = '')
```

## Introduction

@Pintar.etal2018 conducted a field mesocosm experiment to explore the relative effects of predation risk and nutrient availability (a pdf of their publication is included in the GitHub repository). The predation risk manipulation was accomplished by adding either 0, 1, or 2, Golden Topminnows (*Fundulus chrysotus*) to each mesocosm. These small fish are surface-feeding predators and are known to elicit colonization and oviposition avoidance among dytiscid beetles and *Culex* mosquitoes, but not in some hydrophilid beetles. Nutrient-level treatments were created by the addition of 0, 4, or 8 g of rabbit chow (Small World Rabbit Food, Mannapro, St. Louis, MO, 40% protein).

## Methods

Please see @Pintar.etal2018 for greater detail on experimental design and data collection. In brief, the experiment was set up as a 3 X 3 factorial treatment design (9 treatment combinations). The experiment was replicated in 3 separate locations at 3 different times for a total of 9 replicates. This should have resulted in 81 observations total, however some fish died during the experiment rendering those observations not being included in the data resulting in only 69 observations.

```{r }
#| label: dataImport

df1 <- read_csv("Pintar_etal_2018.csv",
                show_col_types = FALSE)

# Block = 1 to 9 for the 9 spatiotemporal blocks (3 locations X 3 times)
# Pool = 1 to 9 for the plastic wading pools used for the experiment
# Location = 1 to 3 for the 3 fields where the pools were deployed
# Time = a to c for the 3 time periods used for each 2-week run of the exp.
# Nutrient = 1 to 3 for the 3 nutrient levels
# Fish = 1 to 3 for the 3 fish densities
# Culex_Egg_Rafts = number of egg rafts collected over the 2 weeks
# Hydrophilidae = number of hydrophilid beetles (Water Scavenger Beetles)
# Dytiscidae = number of dytiscid beetles (Predaceous Diving Beetles)

# Fish and Nutrient are provided as integers, but I wish to use them as ordered
# factors so that, by default, polynomial contrasts will be used to compute the
# ANOVAs.

df1$Nutrient <- factor(df1$Nutrient,
                       levels = c("0", "4", "8"),
                       ordered = TRUE)

df1$Fish <- factor(df1$Fish,
                   levels = c("0", "1", "2"),
                   ordered = TRUE)

df1$Block <- factor(df1$Block)

```

### Statistical Methods

In this statistical analysis, the data was represented by 9 spacio-temporal blocks (3 locations x 3 unique times when the data was collected) each represented by the "Block" row in the dataset. A linear model for multiple variables was created for each dependent variable using the lmer() function from the @lmerTest package. An ANOVA model was created for each variable using the anova() function found in the base R package, and the model was visualized on a table using the kbl() function found in the @kableExtra package. A table for random effects was also created using the rand() function found in the @lmerTest package, and was visualized on a table using the kbl() function.

## Results

@tbl-cx-1 shows us the results of a fixed effect model for the *Culex* egg rafts, where the interactions between *Culex* egg rafts and the nutrient treatment (p = 0.043) as well as the interaction between egg rafts and the fish treatment (p = 0.002) are both statistically significant. The F-statistic for the interaction of egg rafts and fish is larger and the p-value is smaller, suggesting that the relation between the two is stronger. @tbl-cx-2 displays relatively high AIC values for both tests (p = 0.002), but it appears that treating the blocks as fixed may be a slightly better fit for the data. @tbl-hydro-1 shows the results of a fixed effect model for *hydrophilidae*, and none of the interactions were statistically significant. @tbl-hydro-2 displays relatively high AIC values for both tests (p \< 0.05), but it appears that treating the blocks as fixed may be a slightly better fit for the data. @tbl-dy-1 shows the results of a fixed effect model for *dytiscidae*, where the interactions between *dy*. and nutrient treatment (p = 0.026) as well as the interaction between *dy*. and fish treatment (p \< 0.05) are both statistically significant. The F-statistic for the interaction between *dy*. and fish is larger and the p-value is smaller, suggesting that the relation between the two is stronger. @tbl-dy-2 displays the relatively high AIC values for both tests (p \< 0.05), but it appears that treating the blocks as fixed may be a slightly better fit for the data.

```{r}
#| label: tbl-cx
#| tbl-cap: "Linear mixed-effects model results for the dependent variable 'culex egg rafts' from data provided by @Pintar.etal2018."
#| tbl-subcap:
#|     - "Fixed effect."
#|     - "Random effect."
#| layout-nrows: 2

#Run 3 separate tests for each dependent variable.
#Run a linear model for culex egg rafts using the lmer function. Include (1|Block) to account for the block separation.
  #lmer is specifically for linear mixed effect modeling. We use this since we're treating our blocks as a random factor.
culexmod <- lmer(Culex_Egg_Rafts ~ Nutrient*Fish + (1|Block), data = df1)

#Create an anova model for the dependent variable culex egg rafts, then output it to an Anova table to visualize the data.
CulexAnova <- anova(culexmod)

#Now I made a table for the ANOVA data (fixed effects), and a second table to represent the results for random effects.
kbl(
  CulexAnova,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

kbl(
  rand(culexmod),
  col.names = c(
    "Parameters",
    "log-likelihood",
    "AIC",
    "LRT",
    "df",
    "P"
    ),
  digits = c(0,3,3,3,0,4)
  )

#Keep in mind, two df (degrees of freedom) are output, because df1 shows the df for the # of parameters, and df2 shows the df for the residuals.
```

```{r}
#| label: tbl-hydro
#| tbl-cap: "Linear mixed-effects model results for the dependent variable 'hydrophilidae' from data provided by @Pintar.etal2018."
#| tbl-subcap:
#|     - "Fixed effect."
#|     - "Random effect."
#| layout-nrows: 2

#In this code chunk I repeated the process in tbl-cx, but did it with the dependent variable hydrophilidae.
hydromod <- lmer(Hydrophilidae ~ Nutrient*Fish + (1|Block), data = df1)

HydroAnova <- anova(hydromod)

kbl(
  HydroAnova,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

kbl(
  rand(hydromod),
  col.names = c(
    "Parameters",
    "log-likelihood",
    "AIC",
    "LRT",
    "df",
    "P"
    ),
  digits = c(0,3,3,3,0,4)
  )
```

```{r}
#| label: tbl-dy
#| tbl-cap: "Linear mixed-effects model results for the dependent variable 'dytiscidae' from data provided by @Pintar.etal2018."
#| tbl-subcap:
#|     - "Fixed effect."
#|     - "Random effect."
#| layout-nrows: 2

#In this code chunk I repeated the process in tbl-cx, but did it with the dependent variable dytiscidae.
dymod <- lmer(Dytiscidae ~ Nutrient*Fish + (1|Block), data = df1)

DyAnova <- anova(dymod)

kbl(
  DyAnova,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

kbl(
  rand(dymod),
  col.names = c(
    "Parameters",
    "log-likelihood",
    "AIC",
    "LRT",
    "df",
    "P"
    ),
  digits = c(0,3,3,3,0,4)
  )
```

## References
